Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-21 paper/ms.tex
cleaning up
reverting title [ci skip]
some more discussion of astero example

\title{%
Fast and scalable Gaussian process modeling with applications
to astronomical time series and power spectrum estimation
\vspace{-3\baselineskip}
}

@@

All of the following examples show time domain datasets with a clear bias in
favor of large homogeneous photometric surveys but these methods can similarly
be applied to spectroscopy, where wavelength~--~instead of time~--~is the
independent coordinate and other one-dimensional domains.
It is also possible to cast some two-dimensional problems in the \celerite\
framework (\dfmtodo{CITE Ian C}).

@@

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
many of the upcoming large scale photometric surveys \citep{Campante:2016,
Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical parameters
(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
function in the time domain scale poorly to state-of-the-art datasets
\citep{Brewer:2009,Corsaro:2014}.

@@

To demonstrate this method, we will use a very simple heuristic model based on
\dfmtodo{CITE} where the model PSD is given by a mixture of 8 components with
amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
like $\log g$.

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890 observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we will infer
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ mission.
For KIC~11615890, the published asteroseismic parameters based on several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}
We randomly select a month-long segment of \kepler\ data, initialize our
\celerite\ model using a grid search in the parameter space, and then use
\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
the full set of parameters.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
\dfmtodo{Add some more discussion here.}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on the full \kepler\ dataset \dfmtodo{CITE}.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
\caption{A comparison between the empirical PSD and the posterior inference of
the PSD as a mixture of stochastically-driven simple harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission.
    The light gray curve shows the raw periodogram and the black curve has
    been smoothed with a Gaussian filter with \dfmtodo{SOME WIDTH}.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

% DFM: I'm going to redistribute the following in other sections.

% Stars are variable, for better or worse:  their variability provides information about the properties
% of stars, but also results in noise that must be accounted for in detecting and characterizing their
% planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
% goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
% properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
% flux variability and spectral variations.

% A promising technique for addressing both of these goals is the Gaussian Process formalism
% \citep{Rasmussen2006,Gibson:2012}.  The basis of this method is to assume that stellar variability
% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
% correlations is completely specified by a correlation matrix.  In its simplest form, the
% correlation matrix is modeled with a time-independent autocorrelation function (although
% the time-independence can be relaxed), including a white-noise component that may vary
% with each exposure, and the autocorrelation function is described by a functional form that may be
% simply parameterized by a relatively small number of parameters, sometimes referred to as
% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
% describe the autocorrelation function, thus inferring properties of the stellar variabilty
% which may be used to characterize properties of the star, such as the rotation period,
% the asteroseismic frequencies, or granulation noise known as `flicker'
% \citep{Aerts2010,Noyes:1984,Bastien:2013}.
% The goals of planet detection include measuring the Doppler shift of the star, measuring
% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
% look for dynamical interactions between planets.  To properly account for the impact of
% stellar variability on the planet parameters requires treating correlations in the data, and
% to do so also requires evaluating the correlation matrix.

% The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
% $N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
% prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
% its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
% = N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
% using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
% {\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
% $N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
% cases.

% One promising solution is to use approximate techiques based upon the fact that correlation
% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
% solver is only approximate \citep{Ambikasaran:2013,Ambikasaran:2016}.  In addition, the
% solver is complicated and can be challenging to use with general choices of kernel.  Another
% promising approximate technique is to use a uniform resampling of the dataset, and then
% rely on the structure of Toeplitz matrices, which require uniformly space time series,
% to obtain approximate solutions of the likelihood function \citep{Wilson:2015},
% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
% this approach has the drawback of being approximate.  A third approach is to use a kernel
% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
% model \citep[aka \texttt{CARMA};][]{Kelly:2014}.  This approach can model a stationary power-spectrum that
% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
% of the Lorentzians are required to have a specific relation.  In addition, these
% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

% The approach we explore in this paper is based upon a method developed by Press \&
% Rybicki twenty years ago \citep{Rybicki:1995}.  They observed that a correlation
% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
% They also indicated that a combination of two exponential kernels would have a simliar scaling,
% but they did not demonstrate how to extend this to more than two kernel components.

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
% Rybicki-Press' (\texttt{celerite}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{celerite} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{celerite} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

% We show that the extended matrix that embeds this multi-component kernel can be written
% in real notation, which speeds up the evaluation of the likelihood function,
% and allows optimization software to use automatic differential in computing
% derivatives of the likelihood function.

@@

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat

>>>>

\title{%
Fast and scalable Gaussian process modeling with applications
to astronomical time series
\vspace{-3\baselineskip}
}

@@

All of the following examples show time domain datasets with a clear bias in
favor of large homogeneous photometric surveys but these methods can similarly
be applied to spectroscopy, where wavelength~--~instead of time~--~is the
independent coordinate and other one-dimensional domains.
It is also possible to cast some two-dimensional problems in the \celerite\
framework \citep[see][for a potential application]{Czekala:2017}.

@@

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
driver for many of the upcoming large scale photometric surveys
\citep{Campante:2016, Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical
parameters~--~instead relying on bootstrapped uncertainty estimators
\citep{Huber:2009}~--~but more sophisticated methods that compute the
likelihood function in the time domain scale poorly to state-of-the-art
datasets \citep{Brewer:2009,Corsaro:2014}.

@@

To demonstrate this method, we will use a very simple heuristic model based on
where the model PSD is given by a mixture of 8 components with
amplitudes and frequencies specified by $\nu_\mathrm{max}$, $\Delta \nu$, and
several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
This model could be easily extended to include small frequency splitting and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by physical parameters
like $\log g$.

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890 observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we will infer
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ mission.
For KIC~11615890, the published asteroseismic parameters based on several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}
We randomly select a month-long segment of \kepler\ data, initialize our
\celerite\ model using a grid search in the parameter space, and then use
\emcee\ \citep{Foreman-Mackey:2013} to sample the joint posterior density for
the full set of parameters.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.

This model requires about a minute of computation time to find the maximum
likelihood parameters and then it takes about an hour to run the MCMC sampling
to convergence on a dual-CPU laptop.
This is more computationally intensive than traditional methods of measuring
asteroseismic oscillations but is is also orders of magnitude cheaper than the
same analysis using another GP solver.
An in depth discussion of the benefits of rigorous probabilistic inference of
asteroseismic parameters in the time domain is beyond the scope of this paper
but we hope to revisit this opportunity in the future.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/astero-11615890-numax_deltanu_corner.pdf}
\caption{The probabilistic constraints on $\nu_\mathrm{max}$ and $\Delta \nu$
    from the inference shown in \Figure{astero} compared to the published
    value (error bar) based on several years of \kepler\ observations
    \citep{Pinsonneault:2014}.
    \figurelabel{astero-corner}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/astero-11615890-comparisons.pdf}
\caption{A comparison between the Lomb-Scargle estimator of the PSD and the
posterior inference of the PSD as a mixture of stochastically-driven simple
harmonic oscillators.
(top) The periodogram of the \kepler\ light curve for KIC~11615890 computed
    on the full four year baseline of the mission.
(middle) The same periodogram computed using about a month of data.
(bottom) The power spectrum inferred using the mixture of SHOs model described
    in the text and only one month of \kepler\ data.
    The black line shows the median of posterior PSD and the gray contours
    show the 68\% credible region.
    \figurelabel{astero}}
\end{center}
\end{figure}

@@

@@

The \celerite\ formalism also allows a principled approach to the estimation
of the power spectrum of a time series, as well as its uncertainties.
Fourier transforms work for uniformly sampled data;  astronomical data are
rarely (if ever) uniformly sampled.  Periodograms are affected by the window
function and aliasing in datasets, and make estimating the significance of
spectral features ambiguous.  The Bayesian approach to the estimation of a power
spectrum (CITE) avoids these difficulties:  it allows the computation of
uncertainties, it accounts for the window function on the uncertainties of
the power spectrum, and in principle it can treat
\agoltodo{Finish this\ldots}
